Modeling defoliation as a proxy for tree health: A case study using machine-learning and hyperspectral remote sensing data

### 2019-06-03 analysis/paper/Schratz_hyperspectral.tex
paper: section: Dealing with high-dimensionality

\section{Data and study area}
\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain (\autoref{fig:study_area}).
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} spanning a range from 0\% - 100\%.
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{Iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{Ganley2009}.

@@

\subsection{Derivation of indices}
% TODO: Mention different feature sets
\noindent To use the full information from the hyperspectral data, we calculated all possible vegetation indices supported by the \textit{hsdar} package (90 in total) as well as all possible \ac{NRI} combinations.
NRIs were calculated using the following formula:

@@

\noindent To account for geometric offsets (which were reported with up to 1 m from \ac{ICGC}), we used a buffer of two meters around the centroid of the respective tree.
The mean value of all pixels touched by the buffer was assigned as the final value of each index.
In total, $\frac{125*126}{2} = 7875$ NRIs were calculated .
Due to four corrupted bands and numerical problems for some band combinations, some indices returned \texttt{NA} for specific observations.
We removed all indices from the dataset that showed one or more \texttt{NA} values (across all plots) since we valued a single observation more than having an additional NRI as a predictor variable.
In total, 7471 indices had no \texttt{NA} values and were subsequently used as predictors.

@@

\noindent Three algorithms (\textit{xgboost}, \textit{\ac{SVM}} and \textit{\ac{RR}}) were benchmarked on their predictive performance.
Besides the well-known \ac{SVM} algorithm \citep{vapnik1998}, we also used \textit{xgboost} which is ensemble method relying on the idea of tree boosting that gained a lot of attention in recent years \citep{chen2016}.
We added penalized L2 (Ridge) regression to the portfolio due to its ability to handle highly correlated covariates \citep{hoerl1970}.
One of the most popular machine-learning algorithm, Random Forest \citep{breiman2001}, was not considered for this study: Due to the high number of variables, model fitting times in the range hours for a single model fit were not practicable for this work.
These high fitting times are caused by hyperparameter \texttt{mtry} which scales with the number of variables \citep{probst2018}.
After the selection of the best model, we checked if the winning algorithm can achieve a similar performance when using only the most important variables compared to using all variables.
A successful feature selection simplifies the spatial prediction task because the prediction dataset needs to consist of less variables.
Furthermore, model complexity and fitting times are reduced.

@@

\bibliography{Biblio_hyperspectral}

>>>>

\section{Data and study area}
\noindent Airborne hyperspectral data with a spatial resolution of one meter and 126 spectral bands was available for four Monterey Pine (\textit{Pinus radiata}) plantations in northern Spain (\autoref{fig:study_area}).
The trees in the study area plots suffer from infections of invasive pathogens such as \textit{Diplodia sapinea}, \textit{Fusarium circinatum}, \textit{Armillaria mellea} or \textit{Heterobasidion annosum} leading to a spread of cankers or defoliation \citep{mesanza2016, iturritxa2017}.
In-situ measurements of defoliation at trees (as a proxy for tree health) were collected to serve as the response variable \textit{defoliation} spanning a range from 0\% - 100\%.
The fungi are assumed to infect the trees through open wounds, possibly caused by previous hail damage \citep{iturritxa2014}.
The dieback of these trees, which are mainly used as timber, causes high economic damages \citep{ganley2009}.

@@

\subsection{Derivation of indices}
% TODO: Mention different feature sets
\noindent To use the full information from the hyperspectral data, all possible vegetation indices supported by the \textit{hsdar} package (90 in total) as well as all possible \ac{NRI} combinations were calculated.
The following formula was used for the NRI calculation:

@@

\noindent To account for geometric offsets (which were reported with up to 1 m from \ac{ICGC}), a buffer of two meters around the centroid of the respective tree was used.
The mean of all pixels touched by this buffer was calculated and assigned to the observations.
In total, $\frac{125*126}{2} = 7875$ NRIs were calculated .
Due to four corrupted bands and numerical calculation errors for some band combinations, a total of 7471 indices was available for each observation.
If an index returned \texttt{NA} for one plot, the index was removed from all plots.
This decision was based on the fact that due to the mass of available indices and the low number of observations, the latter was valued higher.

\subsection{Benchmarking design}

The benchmarking matrix of this study consists of the following algorithms:

\begin{itemize}
	\item  Extreme Gradient Boosting (\ac{XGBOOST})
	\item  Random Forest (\ac{RF})
	\item  Penalized Regression (both L1 and L2)
	\item  Support Vector Machine (\ac{SVM})
\end{itemize}

\ac{RF} and {SVM} are well established algorithms that are widely used in environmental modeling.
Extreme Gradient Boosting (commonly referred to as \ac{XGBOOST}) showed promising results in benchmark competitions in recent years.
Penalized Regression is a statistical modeling technique that is capable of dealing with highly-correlated covariates by applying a penalization term which shrinks the coefficients of the model.
The penalties are also known as LASSO (L1) and RIDGE (L2).
The former does not allow the removal of variables from the model (penalization to zero) while the latter does.
Both penalties can also be combined.
The combined approach is called "elastic net".

Three feature sets were used in this study:

\begin{itemize}
	\item The raw hyperspectal band information (HR, $n = 122$)
	\item Vegetation Indices (\ac{VI}, $n = 90$)
	\item Normalized Ratio Indices (\ac{NRI}, $n = 7471$)
\end{itemize}

An exhaustive hyperparameter tuning was applied during nested spatial \ac{CV} for all algorithms.
Maximum Bayesian Optimization \citep{mlrmbo} was used for parameter optimization.
This approach first composes \textit{n} randomly chosen hyperparameter settings out of a user defined search space.
After these \textit{n} tries have been evaluated, a new hyperparameter setting, which is going to be evaluated next, is proposed based on a fitted regression model.
The regression model estimates the performance of the machine-learning method for unknown hyperparameter settings.
Using these estimates, a new promising hyperparameter setting is proposed to be evaluated next.
This strategy continues until a termination criterion, defined by the user, is reached \citep{hutter2011, jones1998}.
An initial design of 30 randomly composed hyperparameter settings and a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is the substantial reduction of the tuning budget required to find a setting which is close to the global minimum
This applies when being compared to methods that do not use information from previous runs, such as random search or grid search \citep{bergstra2012}.

A spatial block nested cross-validation was chosen to reduce the influence of spatial autocorrelation as much as possible.
Each tree plot served as one fold within the resampling setup, resulting in four folds total.
For the inner level, $p - 1$ folds were used (with $p$ being the number of plots).

\subsection{Dealing with high-dimensionality}
% TODO talk about filter vs wrapper vs PCA
The case of high-dimensionality in machine-learning is a research field on its own and numerous approaches were developed in recent years to optimize this case.
% TODO talk more about the general problem and solution, maybe 1-2 sentences

So called "Filter methods" are one approach.
The basic concept relies on the ranking of all features of the dataset using certain heuristics.
Various filter methods exist with different capabilities what feature types are supported (numeric, nominal).
After the coveriates have been ranked by a feature method, a decision needs to be made which features to keep and which to discard.
This step is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the goal is to find the best threshold at which the model achieves a good performance while requiring the least amount of variables.

Another approach is to use algorithms that are normally used for hyperparameter optimization such as \textit{Random Search} or a \textit{Genetic Algorithm}.
Here, first a subset of features is chosen based on the selected algorithm.
Now the model is fitted on the data and the performance is evaluated.
The hyperparameter tuning can only start after the feature-selection optimization has ended.
This method is very expensive since two optimization stages need to be run.
This approach is also referred to as "wrapper method" and was not considered in this work due to the expensive runtime.

A third commonly used method is the "Principal Component Analysis".
Here, the main components of the feature space are extracted and combined.
Usually the first two components are taken since these contain the major information of the coveriates.
By using the automatically estimated explained variance of the main components, the model can rely on a few features containing the condensed information of most variables of the data.
This enables the fitting of very cheap models with a minimal loss of predictor information.

% TODO mention that filter run before hyperpar tuning and are cached
% TODO mention that the % value of the remaining features is tuned as a normal hyperpar

@@

We added penalized L2 (Ridge) regression to the portfolio due to its ability to handle highly correlated covariates \citep{hoerl1970}.
These high fitting times are caused by hyperparameter \texttt{mtry} which scales with the number of variables \citep{probst2018}.
A successful feature selection simplifies the spatial prediction task because the prediction dataset needs to consist of less variables.
Furthermore, model complexity and fitting times are reduced.

@@

\bibliography{Biblio}
